A point-level trend model for the North American Breeding Bird Survey that corrects for detectability

### 2024-08-25 article/article.tex
work on introduction

\section{Introduction}
\par Data for the North American Breeding Bird Survey (BBS) have been collected since the 1960s \citep{hudson_role_2017, sauer_first_2017}, and the protocol has remained essentially the same since. 
Each year, observers survey their BBS ``route", which consists of 50, 3-minute point counts spaced roughly 800 m apart from each other. 
The observer is instructed to record every bird they see or hear within a 400 m radius of each point, during each of the 3-minute point counts. 
Counts for each species are summed across the 50 point counts along a route.
Models for BBS data most often rely on the route-level summary counts, because the survey was designed based on routes as the unit of replication and until the mid-1990s, the database did not include the point-level counts.
These data are then modelled as follows:

@@

where $N_t$ is the index of abundance in given year and stratum.
These trends have been used as the gold standard of trend estimates for North American bird populations for decades, and have informed numerous important reports such as the State of Canada's Birds \citep{north_american_bird_conservation_initiative_canada_state_2019} and State of North American Birds \citep{north_american_bird_conservation_initiative_canada_state_2022}, and have been used to estimate the overall loss of close to three billion birds since the 1970s \citep{rosenberg_decline_2019}.

\par TO UPDATE Unmodeled or uncontrolled changing detectability of birds through time would bias estimates of trends from BBS models.
The BBS field methods control for many factors that are known to affect detectability ().
Some sources of potential changing detectability can be modeled using route-level summaries, including changes in traffic during surveys and changes in phenology().
However many of the factors affecting detectability may vary among the points within a given route.
For example, changes in forest cover…

\par In this paper, we describe an extension of the current BBS status and trend model (i.e., Equation \ref{bbs}) that considers BBS data at the point-level. 
Using point-level (also called stop-level) data, we compare trends and trajectories of Ovenbird (\textit{Seiurus aurocapilla}) obtained through this point-level model to trends and trajectories obtained through the current route-level BBS models described in \citet{smith_spatially_2023}.
After establishing this baseline comparison, we incorporate detectability offsets into the new point-level model in two ways: first using a naive approach that considers these detectability offsets as true point values, and secondly by incorporating error associated with the derived detectability offsets and propagating forward the uncertainty using methods described in \citet{bravington_variance_2021}.
Finally, we investigate how incorporating detectability as offsets in both of these ways affects the trends and trajectories for Ovenbird across Canada, and in particular focus on areas where there have been large landcover changes.

>>>>

\section{Introduction}

\par Long-term biological data are vital for assessing status and trends of animal populations through time.
In North America, trends from the North American Breeding Bird Survey (BBS) \citep{sauer_first_2017, hudson_role_2017} have been used as the gold standard of trend estimates for North American bird populations for decades, and have informed numerous important reports such as the State of Canada's Birds \citep{north_american_bird_conservation_initiative_canada_state_2019} and State of North American Birds \citep{north_american_bird_conservation_initiative_canada_state_2022}.
These trends have also been used to estimate the overall loss of close to three billion birds since the 1970s \citep{rosenberg_decline_2019}.
In Canada, trends from the BBS inform the listing and delisting process of species at risk through the Committee on the Status of Endangered Wildlife in Canada.
As such, it is crucial to have precise and accurate trend estimates through time to properly assess how bird populations are doing in time and in space.

\par NEEDS UDPATED. All biological data, including long-term biological data, have challenges. 
One of the principal challenges in any biological data collection is detectability \citep{bennett_how_2024}.
\par In the BBS, unmodeled or uncontrolled changing of detectability of birds through time would bias estimates of trends from BBS models.
The BBS field methods control for many factors that are known to affect detectability ().
Some sources of potential changing detectability can be modeled using route-level summaries, including changes in traffic during surveys and changes in phenology \citep{griffith_traffic_2010}.
However many of the factors affecting detectability may vary among the points within a given route.
For example, changes in forest cover...

\par In this paper, we describe an extension of the current BBS status and trend model (i.e., Equation \ref{bbs}) that considers BBS data at the point-level, to allow for detectability to be explicitly considered. 
Using point-level (also called stop-level) data, we compare trends and trajectories of Ovenbird (\textit{Seiurus aurocapilla}) obtained through this point-level model to trends and trajectories obtained through the current route-level BBS models described in \citet{smith_spatially_2023}.
After establishing this baseline comparison, we incorporate detectability offsets into the new point-level model in two ways: first using a naive approach that considers these detectability offsets as true point values, and secondly by incorporating error associated with the derived detectability offsets and propagating forward the uncertainty using methods described in \citet{bravington_variance_2021}.
Finally, we investigate how incorporating detectability as offsets in both of these ways affects the trends and trajectories for Ovenbird across Canada, and in particular focus on areas where there have been large landcover changes.

\subsection{Current BBS Protocol and Analysis}

\par Data for the North American Breeding Bird Survey (BBS) have been collected since the 1960s \citep{hudson_role_2017, sauer_first_2017}, and the protocol has remained essentially the same since. 
Each year, observers survey their BBS ``route", which consists of 50, 3-minute point counts spaced roughly 800 m apart from each other. 
The observer is instructed to record every bird they see or hear within a 400 m radius of each point, during each of the 3-minute point counts. 
Counts for each species are summed across the 50 point counts along a route.
Models for BBS data most often rely on the route-level summary counts, because the survey was designed based on routes as the unit of replication and until the mid-1990s, the database did not include the point-level counts.
These data are then modelled as follows:

@@

where $N_t$ is the index of abundance in given year and stratum.
